---
title: "Mapping Assignment"
author: "Michael Kohler"
date: "4/7/2109"
output: html_document
---

# Loading and preparing the data

```{r setup}
library(tidyverse)
library(leaflet)
library(sf)

# Load the data
nativity_1890 <- read_csv("census/nhgis0058_ds27_1890_state.csv")

# Load the shapefiles and de-project to lat/long representations
states_1890 <- read_sf("shapefiles/simplified/US_state_1890.shp") %>% 
  st_transform(4326) %>% 
  filter(STATENAM != "Alaska Territory")

# Get the state centers
centroids_1890 <- states_1890 %>% 
  st_centroid() 
```

Some helper functions.

```{r}
# Plot just the geometries
plot(st_geometry(states_1890))
# Plot a variable
plot(states_1890["SHAPE_AREA"])
```


Examine the corresponding codebook (in this case, `census/nhgis0058_ds27_1890_state_codebook.txt`) to see what the column names mean and make the data more amenable to exploration. In this case, `AUM001` is the code for total population, and `AVL016` is the code for born in Germany. We will use that information to make a more useful table.

```{r}
german <- nativity_1890 %>% 
  select(state = STATE,
         year = YEAR,
         GISJOIN,
         population = AUM001,
         german = AVL016) %>% 
  mutate(german_percent = round(german / population, 3)) %>% 
  arrange(desc(german_percent))
```

# Exploratory analysis

You can make a nice table like so.

```{r}
german %>% 
  top_n(10, german_percent) %>% 
  select(-GISJOIN, -year) %>%  
  mutate(german_percent = german_percent * 100) %>% 
  knitr::kable(format.args = list(big.mark = ","))
```

You can also make exploratory graphics, like this histogram of percentages.

```{r}
ggplot(german, aes(x = german_percent)) +
  geom_histogram(binwidth = 0.01)
```

Or this bar plot of the German population.

```{r}
german %>% 
  arrange(desc(german)) %>% 
  mutate(state = fct_inorder(state)) %>% 
  filter(german > 10000) %>% 
  ggplot(aes(x = state, y = german)) +
  geom_col() +
  coord_flip()
```

## Mapping

## Joining data

Our state data is in two variables. `centroids_1890` has the latitude and longitude of the state centers, while `states_1890` has the polygons for the states. Our census data is in the `german` data frame, or more broadly in the `nativity_1890` data frame. We need to bring these two together with a `left_join()`. Luckily, NHGIS provides the `GISJOIN` column. Note: geometries on the left!

## Points

We will start with centroids since they are easier to map.

```{r}
german_points <- centroids_1890 %>% 
  left_join(german, by = "GISJOIN")
```

We can make a leaflet map with similar synatx to ggplot2.

```{r}
leaflet(german_points) %>% 
  addTiles() %>% 
  addMarkers()
```

Markers are not very interesting. We want to set the radius of the circle to the square root of the population.

```{r}
pop_scale <- function(x, max_radius = 20) {
  x %>% 
    sqrt() %>% 
    scales::rescale_max(to = c(0, max_radius))
}
pop_scale(german_points$german) %>% head()
```


```{r}
leaflet(german_points) %>% 
  addTiles() %>% 
  addCircleMarkers(radius = ~pop_scale(german),
                   label = ~state,
                   popup = ~paste0(state, ": ", german),
                   color = "red")
```

## Polygons

First we need to join the polygons to the German data.

```{r}
german_shapes <- states_1890 %>% 
  left_join(german, by = "GISJOIN")
```

Now we can map the polygons.

```{r}
leaflet(german_shapes) %>% 
  addTiles() %>% 
  addPolygons(label = ~state)
```

When we mapped the German population as points, we needed to scale the levels to pixels. Now we need to go from populations or percentages to colors. Leaflet provides a helper function.

```{r}
german_percent_colors <- colorNumeric("PuRd", domain = german$german_percent)
german_percent_colors(german$german_percent) %>% head()
```

Now we can fill in the map.

```{r}
leaflet(german_shapes) %>% 
  addTiles() %>% 
  addPolygons(fillColor = ~german_percent_colors(german_percent),
              fillOpacity = 1,
              color = "black", weight = 1,
              label = ~state,
              popup = ~paste0(state, ": ", 100 * german_percent, "%")) %>% 
  addLegend("bottomright", pal = german_percent_colors, values = ~german_percent,
    title = "German born",
    labFormat = labelFormat(suffix = "%",
                            transform = function(x) {x * 100}),
    opacity = 1
  )
```


